Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-07-08 MS/ms.Rmd
Fixed some bib problems. Did some fixing of text.

Of course, there are certain types of heterogeneity that transcend different questions or that are simply a 'nuisance' to interpreting effects. In other words, if we compare different types of 'apples' is it still fair to synthesize these studies to understand overall effects? This situation might be analogous to a meta-analysis in comparative physiology that attempts to understand the effects of dietary on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of fish under different temperatures, but in the end, they are still all the same fish species, governed by the same thermodynamic effects on growth. In this situation, we might still be interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. It is not always clear how we should tackle these sources of heterogeneity. If temperature differences are of direct interest, then we should definitely model them. If, however, it is less interesting, than say the diet administered, then it may introduce an unnecessary complication to the interpretation of the effect(s) of interest. Regardless, a diversity of factors *a priori* that are hypothesised or known to impact heterogeneity does not preclude us from also exploring all these important moderators or predictors of variability in effects across studies; however, depending on the effect size used it may just complicate their interpretation or make modelling of all these effects challenging.

@@

As discussed above, it is clear that different types of questions and experimental approaches will mix 'apples and oranges' to varying extents, and indeed, this heterogeneity (or basket of fruit) is often the main interest to comparative biologists. To better understand the different types of effect sizes being used, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled we conducted a literature survey. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contrast-based effects) or the conditions under which traits were measured (i.e., for correlations) across studies? 3) If so, was the variation resulting from treatment differences accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We also recorded whether publication bias and sensitivity analyses were conducted, if data were publicly available, where it was stored, and if so, if it was usable (i.e., in a machine readable format).

@@

All searches were conducted in Scopus between May 16 - 18, 2021. Briefly, we first did a series of pilot searches (detailed in Supplementary Materials) to refine our search query. We searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title and abstract, restricting our search to key comparative physiology journals (detailed in Table X - Supplementary Materials) within the last 6 years. We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative synthesis in the field of comparative physiology, but likely provided a random sample to inform on the questions we sought to address (see below).

@@

## **Opening Up New Opportunities in Comparative Physiology: Expanding the Breadth of Effect Sizes to Deal with Treatment Differences Across Studies**

#### *Brief Introduction to Common Effect Sizes and Sampling Errors that are Useful in Comparative Physiology*

@@

A very common occurrence in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using normal effect size measures (see above) would result in the effect sizes varying in the types of temperatures applied within and across across studies contributing to 'comparability' issues. As such, it is often of interest to standardize effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

#### *Comparing changes in mean physiological rates*

@@

#### *New Effect Size Metrics that Compare Changes in Variability Across Treatments*

@@

where $CV$ is the coefficient of variation defined as $SD / R$. Whether one choose to use SD or CV-based effect size depends on questions at hand (discussed in @Nakagawa2015 and Senior et al. 2020 <!--- sorry I have not added it to bib file - I will do this later once I go through --->)

## **Meta-Analytic Models to Control for and Investigate into Heterogeneous Effects Within and Across Studies**

@@

While heterogenity estimates using $I^2$ measures can be useful for understanding the relative contributions of different factors to effect size variation, prediction intervals might be more appropriate in many cases [@Nakagawa2021c; @Borenstein2019]. Unlike $I^2$, prediction intervals are more holistic and probably more meaningful in the context of meta-analysis. Prediction intervals make use of the standard deviation of the mean instead of the standard error (which creates confidence intervals) [@Borenstein2019].

## **Conclusions and Future Directions**
Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental approaches / designs and environments. While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, obvious drivers of effect size heterogeneity, such as temperature and dosage differences across studies, may not really be of interest. Indeed, comparative physiologists are likely to want an effect size or analytical approach that allows them to remove these 'nuisances'. Here, we provide comparative physiologists with a set of tools to deal with 'nuisance heterogeneity' by creatively 'reinventing' standard effect sizes or using meta-regression models. Of course, nuisance heterogeneity might in of itself be the main interest, in which case standard meta-regression approaches will work sufficiently. However, in many cases, comparative physiologists might simply want to focus on bigger, more interesting biological moderators of effect size direction and magnitude. However, estimating complex meta-regression models with a multitude of fixed effects might result in more challenging model interpretation. As such, a combined approach, where some of the known heterogeneity is dealt with by deriving a variant effect size and associated sampling variance, and other biologically interesting moderators are fit in a meta-regression model, might be the desired. Indeed, this may even be a necessity to simplify the model.

>>>>

Of course, there are certain types of heterogeneity that transcend different questions or that are simply a 'nuisance' to interpreting effects. In other words, if we compare different types of 'apples' is it still fair to synthesize these studies to understand overall effects? This situation might be analogous to a meta-analysis in comparative physiology that attempts to understand the effects of dietary on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of fish under different temperatures, but in the end, they are still all the same fish species, governed by the same thermodynamic effects on growth. In this situation, we might still be interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. It is not always clear how we should tackle these sources of heterogeneity. If temperature differences are of direct interest, then we should definitely model them. If, however, it is less interesting, than say the diet administered, then it may introduce an unnecessary complication to the interpretation of the effect(s) of interest.

@@

As discussed above, different types of questions and experimental approaches will mix 'apples and oranges' to varying extents, and indeed, this heterogeneity (or basket of fruit) is often the main interest to comparative biologists. To better understand the different types of effect sizes being used, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled we conducted a literature survey. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contrast-based effects) or the conditions under which traits were measured (i.e., for correlations) across studies? 3) If so, was the variation resulting from treatment differences accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We also recorded whether publication bias and sensitivity analyses were conducted, if data were publicly available, where it was stored, and if so, if it was usable (i.e., in a machine readable format).

@@

All searches were conducted in Scopus between May 16 - 18, 2021. Briefly, we first did a series of pilot searches (detailed in Supplementary Materials) to refine our search query [See @Foo2021 for an overview of this process]. We searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title and abstract, restricting our search to key comparative physiology journals (detailed in Table X - Supplementary Materials) within the last 6 years. We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative synthesis in the field of comparative physiology, but likely provided a random sample to inform on the questions we sought to address (see below).

@@

## **Opening Up New Opportunities in Comparative Physiology: Expanding the Breadth of Effect Sizes to Deal with "Nuisiance Heterogenity"**

#### *Common Effect Sizes and Sampling Errors that are Useful in Comparative Physiology*

@@

#### *Comparing Changes in Mean Physiological Rates, $Q_{10}$*
A very common occurrence in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using normal effect size measures (see above) would result in the effect sizes varying in the types of temperatures applied within and across across studies contributing to 'comparability' issues. As such, it is often of interest to standardize effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

@@

#### *Formalising Effect Size Metrics that Compare Changes in Variability Across Treatments in the Presence of "Nuisance Heterogenity"*

@@

where $CV$ is the coefficient of variation defined as $SD / R$. Whether one choose to use SD or CV-based effect size depends on questions at hand [discussed in @Nakagawa2015 and @Senior2020].

Our example using $Q_{10}$, a well known effect measure in comparative physiology, shows how common effect size measures can be re-formalised to account for 'nuisance hetreogenity', in this case temperature. We can apply similar approaches to other commonly used effect size metrics. We describe more generally how this can be done in Box 1, applying similar principles to standardise temperature differences for slopes and for standardised mean differences.

## **Meta-Analytic Models to Control for and Investigate Heterogeneous Effects Within and Across Studies**

@@

While heterogenity estimates using $I^2$ measures can be useful for understanding the relative contributions of different factors to effect size variation, prediction intervals might be more appropriate in many cases [@Nakagawa2021c; @Borenstein2019]. Unlike $I^2$, prediction intervals are more holistic and probably more meaningful in the context of meta-analysis. Prediction intervals make use of the standard deviation of the mean instead of the standard error (which is used to calculate confidence intervals) [@Borenstein2019].

## **Conclusions and Future Directions**
Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental approaches / designs and environments. While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, obvious drivers of effect size heterogeneity, such as temperature and dosage differences across studies, may not really be of interest. Indeed, comparative physiologists are likely to want an effect size or analytical approach that allows them to remove these 'nuisances'. Here, we provide comparative physiologists with a set of tools to deal with 'nuisance heterogeneity' by creatively 'reinventing' standard effect sizes or using meta-regression models. Of course, nuisance heterogeneity might in of itself be the main interest, in which case standard meta-regression approaches will work sufficiently. However, in many cases, comparative physiologists might simply want to focus on bigger, more interesting biological drivers of effect size direction and magnitude. However, estimating complex meta-regression models with a multitude of fixed effects might result in more challenging model interpretation. As such, a combined approach, where some of the known heterogeneity is dealt with by deriving a variant effect size and associated sampling variance, and other biologically interesting moderators are fit in a meta-regression model, might be the desired. Indeed, this may even be a necessity to simplify the model.
